Reasoning about a Capability Machine with Local Capabilities

### 2017-06-18 popl-2018/paper.tex
intro

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow correctness on
capability machines. Specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities, and its operational semantics
  (Section~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (Section~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions). It also does not use separate
  per-component stacks but a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel Kripke logical relation for reasoning about programs on the
  capability machine. For our logical relation we prove an analogue of
  the standard fundamental theorem of logical relations --- our theorem
  gives  the most general and powerful formulation of the formal guarantees offered by a
  capability machine (a form of capability
  safety~\citet{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem. First, we model the
  natural continuation-passing-style of assembly code (with return pointers as
  continuations) using a \emph{single} orthogonal closure, rather than
  the earlier used biorthogonal closure. Second, we express the special nature of
  local capabilities using a variant of Dreyer et. al.'s public and private
  future worlds~\citep{Dreyer:jfp12}, which Dreyer et. al.\ used to express well-bracketedness
  guarantees in a lambda calculus. The logical relation and the
  fundamental theorem expressing capability safety
  are presented in Section~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (Section~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

>>>>

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow correctness on
capability machines. Specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities, and its operational semantics
  (Section~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (Section~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions). It also does not use separate
  per-component stacks but a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel Kripke logical relation for reasoning about programs on the
  capability machine. For our logical relation we prove an analogue of
  the standard fundamental theorem of logical relations --- to the
  best of our knowledge, our theorem
  is % by far
  the most general and powerful formulation of the formal guarantees offered by a
  capability machine (a form of capability
  safety~\citet{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem. First, we model the
  natural continuation-passing-style of assembly code (with return pointers as
  continuations) using a \emph{single} orthogonal closure, rather than
  the earlier used biorthogonal closure. Second, we express the special nature of
  local capabilities using a variant of Dreyer et. al.'s public and private
  future worlds~\citep{Dreyer:jfp12}, which Dreyer et. al.\ used to express well-bracketedness
  guarantees in a lambda calculus. The logical relation and the
  fundamental theorem expressing capability safety
  are presented in Section~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (Section~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}
